---
title: "NYPD Shooting Data Project"
author: "Carsen Schulz"
date: "2023-09-27"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
#knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Summary of Data
This data is the historical data of NYPD shootings, from 2006 to 2022. It is organized such that each row is an instance of a shooting, while each column gives information about that shooting. Some important attributes included in the data are:

* Occurrence Date
* Occurrence Time
* Borough
* Precinct
* Statistical Murder Flag (whether the shooting was classified as a murder)
* Perpetrator Characteristics
* Victim Characteristics



```{r importData, echo=TRUE}
url <- "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD"
NYPD_data <- read_csv(url)
```

## Tidy the Data

I did two things to tidy the data:

1. Changed the data type of the date column to a date type instead of a character type.

2. Removed unnecessary columns such as location descriptions (I show that over 90% were N/A), GPS location data, info about borough and precinct, and perpetrator/victim information. 

This data is not useless - there are interesting analyses to be done with it. I simply chose not to use that data for this project.

In a broader scope project, I would love to investigate the geospatial relationships between boroughs, for example.

```{r tidyData, echo=TRUE}
NYPD_data <- NYPD_data %>%
  mutate(OCCUR_DATE = mdy(OCCUR_DATE))

sum(is.na(NYPD_data$LOC_OF_OCCUR_DESC)) / nrow(NYPD_data)
sum(is.na(NYPD_data$LOC_CLASSFCTN_DESC)) / nrow(NYPD_data)

NYPD_data <- NYPD_data %>% select(-LOC_OF_OCCUR_DESC, -LOC_CLASSFCTN_DESC, -JURISDICTION_CODE,
                                  -X_COORD_CD, -Y_COORD_CD, -Latitude, -Longitude, -Lon_Lat,
                                  -PERP_AGE_GROUP, -PERP_SEX, -PERP_RACE, -VIC_AGE_GROUP,
                                  -VIC_SEX, -VIC_RACE, -LOCATION_DESC)

summary(NYPD_data)
```

## Year-by-Year Analysis

I am curious whether the number of shootings in New York has gone up or down since 2006. To analyze this, I will make a bar graph of the number of shootings each year. This will require making a new column for "YEAR".

Additionally, in the same bar graph, I want to see how many of those shootings were classified as murders.

The grey bar shown is the number of shootings in that year, and the black portion of the bar represents how many of those shootings were classified as a murder:

```{r yearly_shootings}
# Create a tibble with the year, count of shootings, and count of murders
yearly_summary <- NYPD_data %>%
  mutate(Year = format(as.Date(OCCUR_DATE), "%Y")) %>%
  group_by(Year) %>%
  summarise(ShootingCount = n(), MurderCount = sum(STATISTICAL_MURDER_FLAG))

# Create a bar chart with both total shootings and murders
ggplot(yearly_summary, aes(x = Year)) +
  geom_bar(aes(y = ShootingCount, fill = "Total Shootings"), stat = "identity", position = "identity") +
  geom_bar(aes(y = MurderCount, fill = "Murders"), stat = "identity", position = "identity") +
  labs(title = "NYPD Shootings/Murders Per Year", x = "Year", y = "Total Count") +
  scale_fill_manual(values = c("Total Shootings" = "grey", "Murders" = "black")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + #Rotate the x-axis labels
  labs(fill = "Legend")  # Change the legend title
```

The comparison of shootings to murders is not particularly interesting. It appears that murders are always a relatively stable fraction of shootings (it appears somewhere between 15-25% of shootings are classified as murders every year.) There is no instance in which one year had a high number of shootings but a low comparative classification of murders, or vice versa.

There are two things I found particularly interesting about this graph:

* 2017, 2018, and 2019 are all very low relative to the surrounding years - why?
  + They are also very near the same year-over-year without much variation which seems odd.
  
* 2020 and 2021 are among the highest years, with 2020 approximately doubling 2019 - but this was the time of COVID-19 lockdowns.
  + Does this mean that the lockdowns had no influence on gun violence?
  
In a broader scope report, I would investigate these findings further.

## Murder Model

For my model, I chose to do a simple logistic regression to predict the likelihood of the next shooting incident resulting in a murder based on whether the previous incident was a murder or not. It considers the binary outcome of "murder" or "no murder" and uses the lag information from the previous incident to make the prediction.

This model is very simplified and is not optimized at all, but it gives a rough walkthrough of this type of prediction:

``` {r murder_model}
# Preprocess the data
NYPD_data <- NYPD_data %>%
  mutate(Murder = as.numeric(STATISTICAL_MURDER_FLAG))  # Convert "STATISTICAL_MURDER_FLAG" to numeric

# Create a time series lag feature
NYPD_data <- NYPD_data %>%
  arrange(OCCUR_DATE) %>%
  mutate(Lag_Murder = lag(Murder))

# Filter out the first incident in each group (no previous incident)
NYPD_data <- NYPD_data %>% filter(!is.na(Lag_Murder))

# Build a logistic regression model
model <- glm(Murder ~ Lag_Murder, data = NYPD_data, family = "binomial")

# Make predictions for the next shooting
new_data <- data.frame(Lag_Murder = 1)  # Assuming 1 indicates a shooting in the last incident
prediction <- predict(model, newdata = new_data, type = "response")

# Display the prediction
cat("Predicted Probability of Murder in the Next Shooting:", prediction, "\n")
```

One way I could optimize this while keeping it fairly simple is by grouping incidents by precinct and borough, so that the model could more accurately predict a murder based on location. This could affect the probability if a certain precinct/borough has a lower murder rate/shooting than another.

## Daytime vs. Nighttime Analysis

The next analysis I want to do will require a new column of data, called DAYTIME, which is a boolean value.

DAYTIME is true if the OCCUR_TIME is between 7AM and 7PM (during the day), and FALSE otherwise (at night).

The pie chart shows the distribution between shootings during the day vs. at night.

This confirms my prediction that more shootings happen at night than during the day.

```{r daytime_analysis_vis}
# Create the DAYTIME column
NYPD_data <- NYPD_data %>%
mutate(DAYTIME = as.POSIXlt(OCCUR_TIME, format = "%H:%M:%S")$hour >= 7 &
as.POSIXlt(OCCUR_TIME, format = "%H:%M:%S")$hour < 19)

# Create summary table to count the occurrences of each DAYTIME value
daytime_summary <- NYPD_data %>%
  group_by(DAYTIME) %>%
  summarise(count = n())

# Create pie chart
ggplot(daytime_summary, aes(x = "", y = count, fill = DAYTIME)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar(theta = "y") +
  scale_fill_manual(values = c("TRUE" = "yellow", "FALSE" = "black"), 
                    labels = c("TRUE" = "During the Day", "FALSE" = "At Night")) +
  theme_void() +
  labs(title = "Distribution of Shooting Incidents based on time of day", fill = "Legend")
```

This made me curious about which hour is the highest, and which is the lowest, so I made a new column for "HOUR".

## Hourly Analysis

I created an hourly summary tibble that counts the number of shootings per each hour of the day.

I then visualized this tibble in a bar chart. It clearly agrees with the pie chart above, as more shootings occur in the nighttime hours.

I confirm that by calculating the min and max of the tibble, and printing that to screen.

```{r hourly_analysis_vis}
# Extract the hour component from OCCUR_TIME
NYPD_data$HOUR <- as.POSIXlt(NYPD_data$OCCUR_TIME, format = "%H:%M:%S")$hour

# Create summary table to count the number of incidents for each hour
hourly_summary <- NYPD_data %>%
  group_by(HOUR) %>%
  summarise(count = n())

# Create bar chart
ggplot(hourly_summary, aes(x = HOUR, y = count)) +
  geom_bar(stat = "identity", fill = "blue") +
  labs(title = "Number of Shooting Incidents by Hour of the Day", x = "Hour of the Day", y = "Number of Incidents") +
  scale_x_continuous(breaks = 0:23)  # Ensure x-axis shows all hours from 0 to 23

# Find the hour with the highest number of shootings (maximum count)
max_hour_count <- hourly_summary %>%
  filter(count == max(count)) %>%
  pull(HOUR)

# Find the hour with the lowest number of shootings (minimum count)
min_hour_count <- hourly_summary %>%
  filter(count == min(count)) %>%
  pull(HOUR)

# Print the results
cat("Hour with the highest number of shootings:", max_hour_count, "with", max(hourly_summary$count), "shootings\n")
cat("Hour with the lowest number of shootings:", min_hour_count, "with", min(hourly_summary$count), "shootings\n")
```

As you can see, the highest number of shootings occurred between 11PM and 12PM, while the lowest number occurred between 9AM and 10AM.

## Bias of Data and Analysis

### Possible Biases

No dataset is perfect, because it reflects an imperfect reality. This dataset, for instance, has many possible biases:

1. Selection Bias: There is no way of knowing that this data represents every shooting in New York, because shootings that went unreported are missing.

2. Sampling Bias in Location: Different boroughs or precincts may have different criteria on recording these shootings, and some areas may be more or less likely to report incidents than others.

3. Missing Data Bias: Approximately 1/3 of the perpetrator data is N/A, as well as over 90% of the location description data. This missing data leads to bias because it is not a perfect representation of reality.

4. Reporting Bias: It is possible that shootings that happen at night might be less likely to be reported, which would lead to bias.

### Mitigations for Data Bias

There is no way to mitigate the first bias, but I trust that this data source is appropriately representative of the reality. For the 2nd and 3rd biases, I purposely did not do any analysis based on location or perpetrator description because so much of it was missing.

I think the 4th bias mentioned mitigates itself, because even with the possible lower likelihood of a nighttime shooting being reported, the amount of nighttime shootings in the data was still significantly more than daytime. These missing reports would have only separated the distribution further.




